\chapter{The Gap Register: Theory Meets Practice}
\label{ch:gap-register}

\epigraph{A capability you have built but cannot see acting is, to the person who needs it, indistinguishable from one you never built.}{}

Chapter~\ref{ch:ecosystem-roadmap} committed DreamLab, in print, to running its own prescriptions on its own systems. This chapter supplies the measurement that makes such a commitment falsifiable. It reports a systematic audit of the four surfaces through which humans and agents actually meet in the VisionClaw ecosystem: the async governance forum, the desktop 3D client, the mixed-reality headset, and voice. This audit is scored against both what the canon promises and what the published research says a collaboration surface must do. The finding, stated plainly before the detail: the substrate is stronger than the interaction. Real cryptographic governance and a designed identity spine sit underneath surfaces on which, today, an operator can watch a count and fire a form, and little else.

\section{How the register was built}

The four surfaces are the ones defined and weighted in Chapter~\ref{ch:four-surfaces}. Each received the same treatment from a twelve-agent \meshterm{analysis mesh}: one agent extracted the canon (what the README, the ADR and PRD corpus, and the 2026-07-03 closeout files actually claim); four surveyed the shipping code, one per surface, recording every capability as \emph{working}, \emph{partial}, \emph{scaffolded}, \emph{absent} or \emph{docs-only} against the governed maturity vocabulary of \texttt{ADR-002}; three researched the human-computer-interaction and multi-agent literature on independent search engines, so no single index shaped the theory; and four adversarial judges, one per surface, ranked each surface's canon and code against that literature. Findings were weighted 40/30/20/10 across forum, desktop, mixed reality and voice, matching the weighting fixed in Chapter~\ref{ch:four-surfaces}: the forum carries the governance load, so it carries the most scrutiny.

Every gap is typed by where the break sits, because the remedy differs by type.

\begin{definitionbox}[title={Three kinds of gap}]
\begin{description}[style=nextline, leftmargin=1.5em]
  \item[Theory-to-canon] Published knowledge our own documents never adopted. The fix is a design position, written down.
  \item[Canon-to-practice] Our documents promise a capability the code does not deliver. The fix is engineering, or an honest retraction of the claim.
  \item[Theory-to-practice] Nobody wrote it down and the code lacks it. The fix is both.
\end{description}
\end{definitionbox}

The register holds 28 gaps: 7 rated critical, 17 major, 4 minor. Those counts are reported here exactly as the judges set them, without rounding or dramatisation. Strengths were recorded with the same discipline, and two of them matter to the argument as much as any gap: relay-enforced separation of who may ask from who may decide, and a cryptographic \texttt{did:nostr} agent identity, both of which sit ahead of common 2026 industry practice, against a landscape in which deployed agent governance and self-disclosure are documented as inconsistent and frequently absent \parencite{AgentIndex2025}.

A note on the honesty of the exercise. The auditor and the audited are the same organisation, and selection bias cuts both ways: an author examining their own system may be harsher than an outsider, or blinder. The register mitigates this by tying every gap to a file and line the reader can open, and by scoring against external literature the authors did not write. It does not eliminate the conflict, which Chapter~\ref{ch:open-questions} names directly. The register is offered as a checkable claim, not a verdict.

\section{Forum: a governance plane for one operator}

The forum is the ecosystem's one production human-in-the-loop mechanism, and its foundations are genuinely good. The Agent Control Surface Protocol (ACSP, Nostr kinds 31400--31405) is wired end to end: an agent publishes a signed \texttt{PanelDefinition} or \texttt{ActionRequest} only after an admin has provisioned it atomically into the \texttt{agent\_registry} (\texttt{ADR-097}), and a human approval is a real signer-backed kind-31403 event, not a UI toggle. The relay is the sole enforcement point, so a non-admin who reaches an Approve button still has the event rejected server-side (\texttt{nip\_handlers.rs:360-368}). That defence-in-depth posture, plus content-addressed decisions aligned with the PROV-AGENT direction of provenance research \parencite{Souza2025ProvAgent}, is the substrate for a real audit trail. What sits on top of it is narrower than the canon advertises.

\begin{table}[h]
\centering
\caption{Forum surface: 10 gaps (2 critical, 7 major, 1 minor)}
\label{tab:gaps-forum}
\begin{tabularx}{\textwidth}{X L{3.2cm} L{1.6cm}}
\toprule
\textbf{Gap} & \textbf{Type} & \textbf{Severity} \\
\midrule
Governance plane is admin-only, excluding ordinary humans & canon-to-practice & Critical \\
\addlinespace
No agent disclosure: agents indistinguishable from humans & theory-to-practice & Critical \\
\addlinespace
Decisions binary; graduated-escalation lifecycle is dead code & canon-to-practice & Major \\
\addlinespace
Forum-client decisions optimistic-only; a rejected decision reads as sent & canon-to-practice & Major \\
\addlinespace
No trust-calibration context at decision time; no decision-audit read API & theory-to-practice & Major \\
\addlinespace
No revoke, appeal or supersede authority for a published decision & theory-to-canon & Major \\
\addlinespace
No design response to approval fatigue / automation complacency & theory-to-canon & Major \\
\addlinespace
No web UI for agent-roster administration & canon-to-practice & Major \\
\addlinespace
Cross-relay federation of governance events absent & canon-to-practice & Major \\
\addlinespace
Silent on multi-agent social influence in threads & theory-to-canon & Minor \\
\bottomrule
\end{tabularx}
\end{table}

The most serious finding is that this is an operator-in-the-loop plane, not a community one. The forum-client \texttt{/governance} route is wrapped in \texttt{AdminGatedGovernance}, which bounces any authenticated non-admin to \texttt{/forums} with a toast before the governance page ever mounts (\texttt{app.rs:1135-1143}), and the relay hard-gates kind-31403 responses to admins. The canon's \texttt{architecture.md} Trust and Gating table explicitly grants members the right to view panels; the shipping code shows them nothing. The flagship chokepoint the canon renders in every scenario as ``Forum $\rightarrow$ Human decides'' is, in the product, ``Forum $\rightarrow$ Admin decides,'' and the distributed-judgment claim collapses to a single-operator approval queue. This directly undercuts the ambient-awareness model the canon leans on, in which coordination depends on shared-workspace visibility \parencite{DourishBellotti1992}.

The second critical gap is disclosure. Agent public keys resolve through the identical display-name cache used for every human member, with no bot badge anywhere in either client (\texttt{governance.rs:134-135}, \texttt{user\_display.rs}). An ordinary reader cannot tell an agent from a human at the point of reading its output or its governance panel. This is the one place the external literature is unanimous: the agentic web requires agents to identify themselves and their authorising principal \parencite{AgenticWebNorms2026}, and deployed practice remains inconsistent enough that disclosure failures have already produced public scandals \parencite{AgentIndex2025}. The canon names agents as distinct \texttt{did:nostr} subjects but requires no visible disclosure, and neither does the code.

The seven major forum gaps compound the two critical ones. The decision model is binary in practice. A fully unit-tested \texttt{DecisionOrchestrator} with delegate, promote and precedent transitions exists in \texttt{nostr-bbs-core}, but it is referenced only inside its own test module, and the live relay projection maps only approve to resolved and reject to rejected, parking any other action in \texttt{under\_review} permanently (\texttt{nip\_handlers.rs:1316-1320}). Both the external consensus on risk-tiered graduated delegation \parencite{Tomasev2026Delegation} and the ecosystem's own richer domain model exceed what ships, and \texttt{architecture.md} documents the unreachable lifecycle as if it were live.

Forum-client decisions are optimistic-only. \texttt{PanelCard} fires a fire-and-forget \texttt{relay.publish()} and flips the UI to ``sent'' the moment local signing succeeds, before any NIP-01 acceptance frame returns (\texttt{governance.rs:206-221}). Because the relay independently rejects non-admin and other invalid events, a user can believe they governed an agent action when nothing was accepted. This false-positive success signal miscalibrates trust exactly where calibration matters most. The sibling BBS client already does this correctly with \texttt{publish\_with\_ack}.

At the decision moment the human is shown title, description and buttons, but no reasoning trace, confidence, risk tier or provenance, and there is no \texttt{GET /api/governance/decisions} read path to review precedent afterwards. A gap the repository documents against itself as Finding 5. Exposing reasoning and disagreement, rather than only a final ask, is what the calibrated-trust literature identifies as the precondition for appropriate reliance \parencite{TrustingTeammates2025}, and reliability does not scale with capability, so per-request evidence is not optional.

Three further major gaps are governance omissions the canon never resolved. There is no revoke, appeal or supersede path: the domain model declares a decision immutable, correctable only by a superseding event, yet neither canon nor code specifies who may publish that event or under what authority, leaving an operator who approves in error with no defined recourse against a settled decision that interruptible-agent research treats as a routine need \parencite{Interruptible2026}. The canon frames governance as an accelerant while offering no safeguard against the accelerant becoming a rubber stamp: a single admin facing an undifferentiated stream of binary panels is the textbook trigger for the approval fatigue and automation complacency the empirical record identifies as the first oversight mechanism to fail \parencite{AutomationBias2012}.

The single most authority-laden act, deciding who may act as a governing agent and revoking them, lives entirely outside the product. None of the nine governance REST endpoints is called from any forum-client page, the admin panel carries Channels, Users, NativePods and DangerZone tabs but no Agents tab (\texttt{admin.rs:108-826}), and operators provision agents through \texttt{curl} and the \texttt{dreamlab.toml} roster, precisely the out-of-band configuration the community-moderation literature warns makes bot authority illegible \parencite{Kuo2026Botender}. Cross-relay federation is absent: \texttt{nostr-bbs-mesh} is scaffold-only with zero agent references and is not a dependency of the relay worker, no 38xxx federation handler exists anywhere, and each community's governance plane is therefore an island against a canon that commits to federated substrates. The minor gap rounds out the register: the canon is silent on the emergent social influence of several agents co-present in one thread, a measurable effect given the live six-agent DreamLab roster \parencite{Song2025SocialGroups}.

\section{Desktop: embodiment that never reaches the screen}

The desktop client is the designated observation layer: humans watch here, they judge in the forum. Its bones are strong. The \texttt{ADR-059} embodiment transport is built end to end, from ingest WebSocket through the \texttt{AgentBeamActor} to a binary broadcast and a mounted client render layer, and real live agent data exists server-side, where \texttt{AgentMonitorActor} polls the agentbox Management API for genuine task statuses. The module docstrings that name the deferred gluon force and give a concrete follow-on design model engineering honesty most observability tooling does not meet. Yet almost nothing reaches the screen.

\begin{table}[h]
\centering
\caption{Desktop surface: 8 gaps (2 critical, 5 major, 1 minor)}
\label{tab:gaps-desktop}
\begin{tabularx}{\textwidth}{X L{3.2cm} L{1.6cm}}
\toprule
\textbf{Gap} & \textbf{Type} & \textbf{Severity} \\
\midrule
Embodiment channel inert: live agent activity never reaches the screen & canon-to-practice & Critical \\
\addlinespace
No steering, interruption or approval affordance in the running client & theory-to-practice & Critical \\
\addlinespace
No client-side ACSP case surface; pending judgment invisible in-scene & theory-to-canon & Major \\
\addlinespace
Agent nodes keyed by \texttt{task\_id}, not \texttt{did:nostr} & canon-to-practice & Major \\
\addlinespace
Fabricated ``MCP Connected'' status miscalibrates trust & canon-to-practice & Major \\
\addlinespace
Push-to-talk globally scoped, not directed at a selected agent & canon-to-practice & Major \\
\addlinespace
No pre-action intent legibility; embodiment shows only the past & theory-to-canon & Major \\
\addlinespace
No swarm-level observability layer versus 2026 AgentOps table stakes & theory-to-practice & Minor \\
\bottomrule
\end{tabularx}
\end{table}

The surface's entire reason to exist is inert. The beam pipeline is wired transport-to-render, but its source-position resolver and the agent-avatar renderer both read \texttt{/api/bots/agents}, whose polling loop starts only inside \texttt{BotsClient::start\_polling()}, reachable only via \texttt{BotsClient::connect()}; this call exists nowhere in the boot path (\texttt{bots\_client.rs:111,144} against \texttt{app\_state.rs:1055}). The resolver therefore always returns false, beams never draw, and the avatar layer additionally discards the real \texttt{.nodes} payload that \texttt{AgentMonitorActor} does produce. Running agents are invisible on screen. This is a live-versus-mock legibility failure rather than a mock-data problem: a dead wire returning an empty array, one missing \texttt{connect()} call away from working, while the 2026-07-03 closeout claims the beam actor is ``wired at boot.'' Continuous passive visibility of activity in the shared space is the bedrock requirement for any oversight surface \parencite{DourishBellotti1992, Epperson2025Debugging}, and the closeout claim should be downgraded to ``transport wired, position source unstarted.''

The second critical gap is that no steering, interruption or approval affordance is reachable in the running client. \texttt{AgentDetailPanel} and \texttt{BotsControlPanel} are complete but imported nowhere, and \texttt{AgentDetailPanel} additionally posts to \texttt{/bots/submit-task} and \texttt{/bots/task-status} routes that do not exist server-side. The mounted command box changes only view and physics settings. Against the dominant 2025--2026 pattern of graduated, interruptible autonomy with checkpoint-reset \parencite{Epperson2025Debugging}, the operator watching agents work cannot pause, redirect or veto anything in-scene; the actual loop is fire a form, then wait. This falls into a hole between forum and desktop, because the canon assigns in-flight steering authority to no surface the operator actually watches.

Five major gaps sharpen the picture. No client-side ACSP case surface exists: the server genuinely produces kind-31402 cases and consumes kind-31403 decisions over the forum relay, but a grep of the client for the governance kinds returns zero hits, so an escalation the server just raised is invisible until the human context-switches to a separate app. In-context escalation is what mixed-initiative theory requires so the operator can act inside the cost-of-interruption window \parencite{Horvitz1999Principles}, and the canon never assigns even an ambient pending-case indicator to the surface the operator is actually watching.

Agent nodes are keyed by ephemeral \texttt{task\_id}, not \texttt{did:nostr} (\texttt{bots\_client.rs:14-39}). Because they carry no key, they cannot be addressed by ACSP nor targeted by voice, and the canonical ``one identity across every boundary'' claim is unrealised precisely on the surface where an operator would select an agent \parencite{AuthenticatedDelegation2025}. The identity spine exists in design and is not the addressing primitive here.

The one health signal on the surface is fabricated. \texttt{get\_status()} sets \texttt{connected = true} unconditionally (\texttt{bots\_client.rs:212}) and never calls the \texttt{test\_connection()} that sits at line 231, so the green ``MCP Connected'' dot polled every three seconds reads healthy while the backend is down. This is a direct violation of the \texttt{ADR-002} maturity-honesty rule and the opposite of the honesty the same codebase models in its gluon-deferral comment; a status decoupled from ground truth drives over-reliance exactly when intervention is needed. The fix is one line.

Push-to-talk is well-built but globally scoped, routing to a generic command pipeline with no concept of a selected agent as target, blocked upstream by the same missing \texttt{did:nostr} keying. The last major gap is a canon omission: embodiment renders only what already happened, so beams show that an agent touched a node but never what an agent is about to do, and with no plan-preview or intent overlay the human can react to consequences without pre-empting them, forgoing the intervention window that intent-legibility practice targets \parencite{Nielsen2025Intent}. A minor gap notes that against the swarm-observability baseline now shipping industry-wide \parencite{AgentOps2024}, the operator gets a count and a fabricated health dot, though the underlying per-task data already exists server-side and is merely unsurfaced.

\section{Mixed reality: presence without identity}

The headset is more built than earlier gap maps implied. A user can enter an immersive session, see the live knowledge graph rendered by the same \texttt{GraphManager} the desktop uses, see live agent nodes fed by real telemetry, watch embodiment beams, and grab and drag generic nodes through a real WebSocket round-trip. The cryptographic identity designed into the wire schema is again ahead of a still-draft industry standard. Past observation, the surface is a one-directional window.

\begin{table}[h]
\centering
\caption{Mixed-reality surface: 6 gaps (2 critical, 3 major, 1 minor)}
\label{tab:gaps-mr}
\begin{tabularx}{\textwidth}{X L{3.2cm} L{1.6cm}}
\toprule
\textbf{Gap} & \textbf{Type} & \textbf{Severity} \\
\midrule
Identity-blind: agent nodes carry no \texttt{did:nostr}, no in-headset actor verifiable & canon-to-practice & Critical \\
\addlinespace
No agent-interaction or governance affordance reachable in MR & theory-to-canon & Critical \\
\addlinespace
Embodiment/copresence theory unrepresented (no avatar, gaze, proxemics) & theory-to-canon & Major \\
\addlinespace
Input layer broken: targeting ray at world-origin, gaze absent & canon-to-practice & Major \\
\addlinespace
Voice PTT-to-actor binding is complete code with zero consumers & canon-to-practice & Major \\
\addlinespace
\texttt{enterVR()} never sets \texttt{isXRMode}; XR rendering runs as desktop & canon-to-practice & Minor \\
\bottomrule
\end{tabularx}
\end{table}

The surface is identity-blind. The \texttt{Agent} struct carries \texttt{id}, name, type, status, position, health and workload, and no public-key or DID field (\texttt{bots\_client.rs:15-38}); the wire schema reserves the identity fields but marks them optional until a later phase. A headset user watching a beam has no cryptographic way to know which actor performed the action, so the canon's central trust primitive is absent exactly where embodiment makes actions visible, and ``the graph IS the audit trail'' does not hold in MR against the disclosure-and-registry direction of provenance research \parencite{Souza2025ProvAgent}. The second critical gap is that no agent-interaction or governance affordance is reachable at all: there is zero import of any command or detail panel anywhere under \texttt{client/src/immersive}, and ACSP has no client dispatcher, so a user can see agents act but cannot inspect, steer, approve or command anything from within the headset. Even a monitoring surface needs a legible intervention path \parencite{Epperson2025Debugging}, and the canon treats XR as pure rendering rather than assigning it one.

The three major gaps follow from those two. Copresence theory is unrepresented: agents render as graph nodes with no avatar, gaze or proxemic semantics, and the multi-user avatar renderer and Vircadia presence bridge exist as code but are never instantiated (\texttt{autoConnect=false}), so the one surface where spatialised copresence is native uses a desktop-2D mental model at room scale, forgoing the trust and legibility that spatial embodiment materially changes \parencite{Schmid2024Objects}. The severity is sharpened by provenance: this is a regression against the author's own quantified 2022 design brief, which fixed the gaze-offset tolerances and turn-passing costs a credible copresence system must respect \parencite{OHare2022Convergence}. These constructs are not merely unmet 2024 standards; they are the operator's own published numbers, unimplemented on the operator's own surface.

The input layer that would support in-headset steering is broken. The agent-targeting ray in \texttt{VRAgentActionScene} never receives a real XR pose and so originates at world-origin, its select callback is never wired through \texttt{VRGraphCanvas}, hand-tracking sets \texttt{isTracking}/\texttt{isPointing} booleans only with a code comment conceding it ``would need XRFrame to get actual pose,'' and gaze is entirely absent. Every downstream step (scoped command, ACSP request, voice binding) is therefore built on a selection that does not work \parencite{SpeakStaySilent2026}.

Voice push-to-talk and the LiveKit service are complete, well-written classes with zero consumers, survivors of a July 2026 dead-code pass that deleted 44 sibling orphans; even wired, they would sit atop the two broken preconditions above. The minor gap is a silent state defect: the working VR entry point never flips \texttt{isXRMode}, so XR-conditional rendering runs in desktop mode throughout a live session, invisible at the documentation layer and surfacing only on-device.

\section{Voice: one closed loop, narrowly scoped}

One genuine end-to-end voice loop exists, and it is honestly the best-behaved thing in the register. \texttt{VoiceInterfaceActor} (\texttt{ADR-110}) subscribes at boot to the Whisper transcription broadcast, gates on a conservative verb-plus-noun grammar, routes to the same settings-assistant LLM the typed control centre uses, and confirms aloud through Kokoro TTS. That is a real human-speaks, agent-hears, agent-responds-by-voice turn, with two strengths that exceed industry practice: identity is asserted from the operator's configured key and never from the transcript (``a transcript cannot assert an identity it does not hold''), and an unrecognised utterance falls back to a read-only query rather than a silent mutation. The loop is scoped to interface tweaks, not agent dialogue or governed action.

\begin{table}[h]
\centering
\caption{Voice surface: 4 gaps (1 critical, 2 major, 1 minor)}
\label{tab:gaps-voice}
\begin{tabularx}{\textwidth}{X L{3.2cm} L{1.6cm}}
\toprule
\textbf{Gap} & \textbf{Type} & \textbf{Severity} \\
\midrule
Canonical PTT voice-to-selected-actor governed-action loop does not exist & canon-to-practice & Critical \\
\addlinespace
No multi-agent addressing or turn-taking model & theory-to-canon & Major \\
\addlinespace
No conversational grounding or repair & theory-to-canon & Major \\
\addlinespace
Docs describe a deprecated voice-to-swarm path as live & canon-to-practice & Minor \\
\bottomrule
\end{tabularx}
\end{table}

The canon documents push-to-talk capturing the selected agent node and dispatching a scoped, signed ACSP request to that agent's \texttt{did:nostr} as step one of its flagship journey. No wired path connects live speech to a selection-aware, signed, addressed agent action. The one producer built for it, agentbox's \texttt{/v1/voice-intent}, is gated off by default (\texttt{agentbox.toml:34}), accepts only a flat optional actor string with no scene-selection binding, and has zero callers in VisionFlow's own STT pipeline; the addressing key it would need does not exist, because agent nodes remain \texttt{task\_id}-keyed \parencite{AuthenticatedDelegation2025}. The canon prose for this step notably lacks the file-and-line evidence that surrounds every genuinely shipped claim elsewhere in the same document.

The two major gaps are theory the canon never adopted. Nothing tells the human which agent is listening: awareness cues are one-directional, two competing swarm-keyword paths fire on the same audio with no floor management, and multi-party addressing has no convention here at all \parencite{SpeakStaySilent2026}. There is no grounding or repair move: unrecognised speech degrades to silence or a read-only default, never a clarifying question, exactly the failure the grounding literature names as central for voice agents, which issue repair requests far less often than humans do \parencite{GroundingRifts2025}. The minor gap is documentation drift: the how-to still describes a voice-to-swarm path that the code has actively deprecated to a canned ``use REST API instead.''

\section{Four surfaces, four recurring patterns}

Read across the surfaces, the 28 gaps resolve into four patterns, and the pattern is more instructive than any single gap.

First, the mesh acts but cannot yet be seen acting. The desktop beam pipeline is complete and inert, the MR surface renders beams without a verifiable actor behind them, and no surface offers the swarm-level observability that is now table stakes. Real agent activity exists server-side on every surface and reaches the operator's eye on none. This is the signature failure of the whole register: capabilities built to the point of one missing wire, then left unstarted.

Second, the identity spine is strong at rest and unused in interaction. The \texttt{did:nostr} identity is designed into the provisioning path, the relay gating and the wire schema, and it is ahead of a still-draft industry standard \parencite{AIIdentity2026}. At the point of interaction it disappears: agent nodes are keyed by \texttt{task\_id} on desktop, on MR and in voice, so the one identity that is meant to cross every trust boundary is never the addressing primitive on any surface where a human would select an agent. Three separate gaps (ACSP addressing, voice targeting and MR verification) share this single root.

\begin{keyinsight}[title={The register's through line}]
The ecosystem's failures are rarely ``it does not work.'' They are ``it is built, and unwired.'' A cryptographic identity that is never read at interaction time; a governance plane real enough to reject a forged approval yet open to only one operator; a beam transport complete to the last actor and starved by an absent \texttt{connect()} call. The distance between the substrate and the surface is the distance this edition asks to be measured on.
\end{keyinsight}

Third, governance is real but admin-only. The forum enforces, at the relay, who may ask and who may decide, an achievement most deployed agent systems lack. It also confines every decision to a single administrator role, with a binary control, no risk-tiering, no disclosure of the deciding actor, and no appeal. Distributed human judgment, the report's own premise, collapses to a single approval queue that the automation-complacency literature predicts will be the first thing to fail.

Fourth, every surface's voice story ends at the same unconsumed binding. Desktop push-to-talk is globally scoped; MR push-to-talk has zero consumers; the voice surface's flagship loop is gated off and uncalled. All three block on the identical pair of missing preconditions, \texttt{did:nostr}-keyed agent nodes and a working selection, so what looks like four independent voice gaps is one unbuilt binding seen from four angles. The type distribution says the same thing another way: of the 28 gaps, the largest class is canon-to-practice, documents promising what the code does not yet deliver. That is a more recoverable position than a system whose theory is wrong, and a more embarrassing one to leave standing in print.

\section{Handing off to remediation}

This chapter measures; it does not prescribe. The remediation sits in Chapter~\ref{ch:ecosystem-roadmap}, and the mapping is close to one-to-one. The roadmap's first priority, closing the governed-writeback security gaps, must precede everything the forum register flags. Surfacing the broker case queue inside the client answers the desktop ``no ACSP surface'' gap and the forum ``admin-only, binary'' gaps together. Threading a \texttt{did:nostr} onto agent nodes is the shared unblocker for the identity-spine pattern across desktop, MR and voice.

Chapter~\ref{ch:open-questions} then takes up the gaps that no engineering ticket closes: the ones where the canon itself is silent and the theory is unsettled, from multi-human decision conflict to XR-specific consent. This chapter's job was narrower and prior to that one. It fixed the numbers. Whether those commitments are met, or conspicuously not met, is what the next edition will report against this register, which is the point of stating the counts, the file paths and the severities exactly, now, before the work is done.

\section{One sprint later}

The audit above is the historical record, and it stands unedited: the counts, the file paths and the severities are what the judges set in July 2026, before any remediation. What follows is a coda written after the gap-close sprint the register triggered, and it reports three things the sprint taught about the register itself.

First, the register had drifted from the code it described. A pre-sprint reconnaissance opened the current tree against the register's own claims and found four items already closed before a line of sprint work began. The ontology \texttt{/propose} endpoint the roadmap flagged as unauthenticated was already gated behind \texttt{RequireAuth::authenticated()}; the \texttt{/load} validation backdoor was already reduced to a single power-user scope; the solid-pod-rs non-destructive PATCH bug had been fixed on 29 May 2026 with three passing regression tests; and the forum's ``missing'' NIP-42 authentication was a deliberate design decision, the edge relay gating by pubkey whitelist with \texttt{auth\_required} set false and NIP-42 answering living only in the browser client. Two further framings were plainly wrong. The desktop embodiment channel the register called ``inert'' in fact awaited \texttt{websocketService.connect()} at boot all along, its residual being live beam traffic rather than a dead wire; and the honest ``MCP Connected'' badge was never the fabricated signal, which was a separate literal-\texttt{true} WebSocket dot. An auditor reading an old snapshot of its own system is the selection bias Chapter~\ref{ch:open-questions} names, seen from the inside. The register was closer to done than it read.

Second, closing the rest was not clean, and the mess is the point. Every closure was checked by a verifier holding a refute mandate, drawn from a different model family than the agent that produced the work, and that verification caught false closures in all three waves and sent them back. Three surfaced in the P0 correctness wave: two genuine, fixed and re-confirmed, plus one fabricated claim in a verifier's own report, caught and discarded. Six surfaced in the P1 measurement wave, each fixed before the wave was allowed to close. Three surfaced in the P2 extension wave, of five first-pass refutations, the other two being stale audits against tiers the implementers had already corrected. One catch survived a full wave: the desktop interrupt affordance was declared closed in P1, and re-verification found it still could not stop an MCP-native agent, because that surface exposes no terminate verb at all. The honest fix was to add a join key where one existed and disclose the non-interruptible state in the client, so an operator sees a control that does not lie about its reach. The auditor and the audited being the same organisation is the standing objection to this whole exercise; the twelve caught false closures are the evidence that, run against itself with a refutation mandate, the discipline has teeth.

Third, the four surfaces moved, unevenly, in the shape the weighting predicts. Scored on the 40/30/20/10 lens of Chapter~\ref{ch:four-surfaces}:

\begin{itemize}
  \item \textbf{Forum (40).} Every ``closed'' condition is met in code: agent badges and read-only member panels, three decision outcomes reachable through the relay projection, a decision-audit read API, relay-confirmed decisions replacing the optimistic-send bug, an in-UI roster admin tab, and the NIP-42 claim reconciled with the mechanism. The heaviest surface is the most closed.
  \item \textbf{Desktop (30).} The beam-and-avatar poll starts at boot, node selection opens steering and approval controls, the status dot reflects real reachability, and agent nodes carry a verified \texttt{did:nostr}. One boundary stands, now visible rather than hidden: an MCP-native external agent cannot be interrupted, and the client says so.
  \item \textbf{Mixed reality (20).} The Godot copresence chain (gaze, proxemics, avatar state, a selection arbiter, an in-headset intervention panel, an amber-until-verified identity badge) is complete and property-tested, and it sits at \texttt{standalone}, not \texttt{integrated}. It cannot reach the higher tier in the build container because there is no headset. This surface is short of its target by hardware, not by code.
  \item \textbf{Voice (10).} The push-to-talk loop is wired end to end from selected actor to signed request to Kokoro acknowledgement, a low-confidence clarification turn exists, and the documentation matches shipped behaviour.
\end{itemize}

The honesty the register demanded of the audit, it now demands of the closure. Nothing above is scored on a live run. No live stack executes in the build container: source bind-mounts resolve on the host rather than the container filesystem, there is no headset, and there is no push credential to fire a GitHub Actions gate. Rather than block three waves on infrastructure the box cannot supply, the loop-closing observations were stacked to a single live session at sprint end, a deliberate relaxation of the wave-gating rule recorded as a deviation rather than smuggled past it. Nine canaries wait on that run: the desktop beam traffic, the \texttt{did:nostr} Schnorr round-trip, the voice end-to-end, the broker case, the relay tap, the forum badge render, two continuous-integration gates that must turn a probe pull request red, and the two in-headset canaries. Each is mechanism-complete and passes its offline test; none has fired against a live stack. Until it does, its item is held at the tier its offline evidence supports, and this chapter's discipline holds: a green canary is the difference between \texttt{integrated} and a claim. The register measured the distance between the substrate and the surface. One sprint later that distance is a single live session wide, and the register says so in those terms, because saying otherwise is the failure it was written to catch.

\section{Forward-stamp: broker governance closes at code}

Between this edition's July snapshot and its going to press, the broker-governance items the coda held open moved --- not to a live-session green, but from design to landed code, on the sibling repositories' \texttt{main} lines. The gap register's own v1.3 addendum records the chain; it is summarised here because three of this chapter's headline gaps are among the items that closed, and a reader owed the honest boundary is also owed the delta.

Five findings closed at code across 10--11 July 2026. The false write-back closure the register flagged as the governance bleed (GOV-1) is fixed: AgentBox's broker-bridge now keys case closure on VisionClaw's real \texttt{writeback\_committed} response and forwards the actual deciding pubkey rather than a placeholder \texttt{'unknown'}, and the \texttt{buildAuthorityGate} check was carried onto the broker decide path itself, blocking a zero-tolerance action until a verified approving \texttt{kind-31403} releases it (agentbox \texttt{a70dc4fb}). The agent-side application gap the closeout phrased as ``routes to a \texttt{BrokerActor} absent from main'' was reframed by VisionClaw \texttt{ADR-130}: there is no distributed \texttt{BrokerActor} by design, so the residual was never to build one but to join the two halves of the loop --- the pending-case producer and the REST decision projection --- which landed so that \texttt{GET /api/broker/inbox} shows open work and each decision projects to the forum as a \texttt{kind-31403} event (GOV-3, \texttt{ca145a1ce}). The git-bridge's \texttt{WriteBackSaga} no longer posts into a 404: \texttt{POST /api/ingest/writeback} is registered with a reachability test (GOV-4, \texttt{78759494e}). And the flagship elevation loop gained the two termini the canon promised: an EL++ consistency gate on the approve arm that fails closed when the drafted axioms are inconsistent or the reasoner cannot be built, opening no PR (GOV-7, \texttt{43a12e401}), and a terminal \texttt{concept\_elevated} event fired on a merged pull request rather than the elevation being claimed at PR-creation (GOV-2, \texttt{ada2069a9}, \texttt{43a12e401}).

One canary fired in test, and its status is worth stating precisely because the discipline this chapter demands turns on the distinction. On 11 July a full REST loop ran against a live server process and a live local relay: \texttt{POST /api/ingest/writeback} returned an attributed, committed approval carrying a PROV-O activity URN, \texttt{GET /api/broker/inbox} showed the case decided, and the \texttt{kind-31403} decision event was read back off the relay by an independent WebSocket subscriber rather than merely client-acknowledged. That is fired-in-test evidence that the loop mechanics --- persist, attribute, commit, project, subscribe --- are sound. It is not the live-session promotion. The run used a local relay and a curl-driven decision, not the production forum relay and a human deciding through the forum UI, and the forum still hard-gates the \texttt{kind-31403} response to an admin key. The broker case end-to-end therefore stays \texttt{pending-live-session}, exactly as the coda's canary list holds it: a live run needs the ACSP panel key registered as an authorised broker pubkey on a running production relay, alongside the VisionClaw server. Two items remain honestly open beyond that boundary --- the \texttt{ShareOrchestratorActor} that would execute a share-transition plan is scaffolded, not built, and GOV-2's merge-poll has yet to observe a real merged PR --- and the stack-up live session that would turn the nine canaries green is still owed. The distance the register measured is unchanged in kind; what changed is that it is now a distance in liveness, not in code.
